Cost- and Deadline-constrained Scheduling of Scientific Workflow Ensembles in IaaS Clouds

### 2012-01-20 paper/paper.tex
Update SPSS description

The simplest strategy for executing ensembles of workflows in the cloud is to use
provision resources statically and schedule workflow tasks dynamically. Given a 
budget $b$, deadline $d$, and the hourly price $p$ of a VM it is easy to calculate
the number of VMs, $N_{VM}$, to provision so that the entire budget is consumed before
the deadline:

@@

The SPDS algorithm statically provisions VMs at the start of the ensemble execution 
and keeps them running until the deadline is reached, or the budget runs out.
This provisioning plan has the advantage that it minimizes the number of provisioning 
and de-provisioning requests.

@@

\subsection{Dynamic Provisioning Dynamic Scheduling (DPDS)}

@@

\subsection{Static Provisioning Static Scheduling (SPSS)}

The previous algorithms are all online algorithms that make provisioning 
and scheduling decisions at runtime. As such, they are not able to analyze 
alternative decisions in order to optimize the schedules.

The SPSS algorithm creates a provisioning and scheduling plan before running 
any workflow tasks.

The approach is to schedule each workflow in the ensemble in priority order, 
rejecting any workflows that cannot be completed by the deadline or that cause 
the cost of the plan to exceed the budget. Once the plan is complete, then the 
workflows are executed according to the plan.

The disadvantage of the static planning approach used by SPSS is that it is 
sensitive to dynamic changes in the environment and the application that may 
disrupt the desired plan.

Algorithm~\ref{alg:admit} shows how ensembles are planned in SPSS.

\begin{algorithm}
\caption{Ensemble planning algorithm for SPSS}
\label{alg:admit}
\begin{algorithmic}[1]
\Require $W$: workflow ensemble; $b$: budget; $d$: deadline
\Ensure Schedule as much of $W$ as possible given $b$ and $d$
\Procedure{PlanEnsemble}{$W,b,d$}
    \State $P\gets empty\ plan$
    \State $A\gets \emptyset$ \Comment{Set of admitted DAGs}
    \For{$w\ in\ W$}
        \State $P^\prime \gets$\ \Call{PlanDAG}{$w,P,d$}
        \If{$P^\prime\ is\ a\ valid\ plan$}
            \If{$Cost(P^\prime) \le b$}
                \State $P\gets\ P^\prime$ \Comment{Accept new plan}
                \State $A \gets A\ +\ w$ \Comment{Admit w}
            \EndIf
        \EndIf
    \EndFor
    \State \textbf{return} $P,A$
\EndProcedure
\end{algorithmic} 
\end{algorithm}

The intuition behind this algorithm is that more workflows can be admitted
if each workflow is able to achieve the lowest-cost schedule that still
meets the ensemble deadline.

In order to create a plan for a single workflow, the SPSS algorithm assigns 
sub-deadlines to each individual task in the workflow, and then schedules 
each task so as to minimize the cost of the task while still meeting its 
assigned sub-deadline. The idea is that if each task can be completed by 
its deadline in the least expensive manner possible, then the cost of the
entire workflow is being minimized without exceeding the ensemble deadline.

SPSS assigns sub-deadlines to each task based on the \emph{float time}
of the workflow. The float time is the amount of extra time that a workflow
can extend its critical path and still be completed by the ensemble deadline.
For a workflow $w$, the float time of $w$ is:
$$
FT(w) = d - CP(w)
$$
where $d$ is the ensemble deadline and $CP(w)$ is the critical path of $w$.
Note that we assume that $CP(w) \leq d$, otherwise the workflow cannot be
completed by the deadline and must be rejected.

SPSS distributes the float time of the workflow by level, so that each level
of the workflow gets a portion of the workflow's float time proportional to
the number of tasks in the level, and the total runtime of tasks in the level.
The idea is that workflows with many tasks and large runtimes should be given
a larger portion of the float time so that those levels may be serialized.
Otherwise, many resources need to be allocated to run all of the tasks in 
parallel, which may be more costly.

@@

where $N(w)$ is the number of tasks in the workflow, $N(l)$ is the number of tasks 
in level $l$, $R(w)$ is the total runtime of all tasks in the workflow, $R(w)$ is
the total runtime of all tasks in level $l$, and $\alpha$ is a parameter between
0 and 1 that causes more float time to be given to levels with more tasks (large 
$\alpha$) or more runtime (small $\alpha$). We set $\alpha$ to be around .7.

The deadline of a task $t$ is then:
$$
DL(t) = LST(t) + RT(t) + FT(Level(t))
$$
where $Level(t)$ is the level of task t, $RT(t)$ is the runtime of $t$, and $LST(t)$ 
is the latest start time of $t$ determined by:
$$
LST(t) = \begin{cases}
0,&$if t is a root task$\cr
\max_{p \in Pred(t)} DL(p),&$otherwise.$\cr
\end{cases}
$$

For each task in the workflow the \textsc{PlanDAG} procedure chooses the 
least expensive scheduling solution that is able to complete the task by 
its deadline. The algorithm allocates VMs in blocks of one billing cycle 
(one hour) regardless of the size of the task. When computing the cost 
of scheduling a task on a given VM it considers idle time from blocks that 
were allocated for previous tasks to be free, while new blocks cost full
price. For example, if a task has a runtime of 10 minutes and the price of
a one-hour billing cycle is \$1, then the algorithm will either schedule 
the task on an existing VM that has 10 minutes of idle time for a cost of 
\$0, or it will allocate another block on an existing or new VM for a cost
of \$1. In addition, the algorithm always prefers to extend the runtime of 
existing VMs rather than allocating new VMs, and if the cost of scheduling 
the task on two different VMs is equal, then the solution with the earliest 
start time is preferred. The result of this is that the algorithm will only
allocate a new block if there are no existing blocks with enough idle time
to complete the task by its deadline, and it will only allocate a new VM if
it cannot add a block to the start or end of an existing VM to complete the
task by its deadline.

\begin{algorithm}
\caption{DAG planning algorithm for SPSS}
\label{alg:plandag}
\begin{algorithmic}[1]
\Require $w$: DAG; $P$: current plan; $d$: deadline
\Ensure Create plan for $w$ that minimizes cost and meets deadline $d$
\Procedure{PlanDAG}{$w,P,d$}
    \State $P^\prime\gets$ copy of $P$
    \State \Call{DeadlineDistribution}{w,d}
    \For{$t\ in\ w\ sorted\ by\ Deadline(t)$}
        \State Choose resource r in $P^\prime$ that minimizes Cost(t)
        \If{$AFT(t,r) < Deadline(t)$}
            \State Schedule(t,r)
        \Else
            \State Provision a new resource r
            \State Schedule(t,r)
        \EndIf
    \EndFor
    \State \textbf{return} $P^\prime$
\EndProcedure
\end{algorithmic} 
\end{algorithm}

An example schedule generated by SPSS is shown in Fig.~\ref{fig:spss-example}.
This example shows how SPSS tends to start many workflows in parallel, running
each workflow on only a few VMs over a longer period of time. In comparison, 
the dynamic algorithms tend to run one workflow at a time across many VMs.

>>>>

The simplest strategy for executing ensembles of workflows in the cloud is to
provision resources statically and schedule workflow tasks dynamically. Given a 
budget $b$, deadline $d$, and the hourly price $p$ of a VM, it is easy to calculate
the number of VMs, $N_{VM}$, to provision so that the entire budget is consumed before
the deadline:

@@

The SPDS algorithm statically provisions $N_{VM}$ VMs at the start of the ensemble execution 
and keeps them running until the deadline is reached, or the budget runs out.
This provisioning plan has the advantage that it minimizes the number of provisioning 
and de-provisioning requests.

@@

\subsection{Dynamic Provisioning Dynamic \\Scheduling (DPDS)}

@@

\subsection{Static Provisioning Static Scheduling \\(SPSS)}

The previous algorithms are all online algorithms that make provisioning 
and scheduling decisions at runtime. In comparison, the SPSS algorithm 
creates a provisioning and scheduling plan before running any workflow tasks.
This enables SPSS to start only those workflows that it knows can be completed
given the deadline and budget constraints, and eliminates any waste that may
be allowed by the dynamic algorithms.

The approach used by SPSS is to plan each workflow in the ensemble in priority 
order, rejecting any workflows that cannot be completed by the deadline or that 
cause the cost of the plan to exceed the budget. Once the plan is complete, then 
the VMs are provisioned and tasks are executed according to the schedules given
by the plan.

The disadvantage of the static planning approach used by SPSS is that it is 
sensitive to dynamic changes in the environment and the application that may 
disrupt the carefully constructed plan.

Algorithm~\ref{alg:admit} shows how ensembles are planned in SPSS. The algorithm
starts with an empty plan containing no VMs and no scheduled tasks. Workflows
from the ensemble are considered in priority order. For each workflow, SPSS
attempts to build on top of the current plan by provisioning VMs to schedule 
the tasks of the workflow so that it finishes before the deadline with the least
possible cost. If the cost of the new plan is less than the budget, then the 
new plan is accepted and the workflow is admitted. If not, then the new plan is 
rejected and the process continues with the next workflow in the ensemble. The 
idea behind this algorithm is that, if each workflow can be completed by the 
deadline with the lowest possible cost, then the number of workflows that can 
be completed within the given budget will be maximized.

\begin{algorithm}
\caption{Ensemble planning algorithm for SPSS}
\label{alg:admit}
\begin{algorithmic}[1]
\Require $W$: workflow ensemble; $b$: budget; $d$: deadline
\Ensure Schedule as much of $W$ as possible given $b$ and $d$
\Procedure{PlanEnsemble}{$W,b,d$}
    \State $P\gets \emptyset$ \Comment{Current plan}
    \State $A\gets \emptyset$ \Comment{Set of admitted DAGs}
    \For{$w\ in\ W$}
        \State $P^\prime \gets$\ \Call{PlanWorkflow}{$w,P,d$}
        \If{$Cost(P^\prime) \le b$}
            \State $P\gets\ P^\prime$ \Comment{Accept new plan}
            \State $A \gets A\ +\ w$ \Comment{Admit w}
        \EndIf
    \EndFor
    \State \textbf{return} $P,A$
\EndProcedure
\end{algorithmic}
\end{algorithm}

In order to plan a workflow, the SPSS algorithm assigns 
sub-deadlines to each individual task in the workflow, and then schedules 
each task so as to minimize the cost of the task while still meeting its 
assigned sub-deadline. The idea is that if each task can be completed by 
its deadline in the least expensive manner possible, then the cost of the
entire workflow can be minimized without exceeding the ensemble deadline.

SPSS assigns sub-deadlines to each task based on the \emph{float time}
of the workflow. The float time is the amount of extra time that a workflow
can extend its critical path and still be completed by the ensemble deadline.
For a workflow $w$, the float time of $w$ is:

$$
FT(w) = d - CP(w)
$$

where $d$ is the ensemble deadline and $CP(w)$ is the critical path of $w$.
We assume that $CP(w) \leq d$, otherwise the workflow cannot be completed 
by the deadline and must be rejected. For large ensembles we expect the 
critical path of any individual workflow to be much less than the deadline.

SPSS distributes the float time of the workflow by level, so that each level
of the workflow gets a portion of the workflow's float time proportional to
the number of tasks in the level and the total runtime of tasks in the level.
The idea is that levels containing many tasks and large runtimes should be given
a larger portion of the float time so that tasks in those levels may be serialized.
Otherwise, many resources need to be allocated to run all of the tasks in 
parallel, which may be more costly.

@@

where $N(w)$ is the number of tasks in the workflow, $N(l)$ is the number of tasks 
in level $l$, $R(w)$ is the total runtime of all tasks in the workflow, $R(w)$ is
the total runtime of all tasks in level $l$, and $\alpha$ is a parameter between
0 and 1 that causes more float time to be given to levels with more tasks (large 
$\alpha$) or more runtime (small $\alpha$). We set $\alpha$ to be 0.7.

The deadline of a task $t$ is then:

$$
DL(t) = LST(t) + RT(t) + FT(Level(t))
$$

where $Level(t)$ is the level of $t$, $RT(t)$ is the runtime of $t$, and $LST(t)$ 
is the latest start time of $t$ determined by:

$$
LST(t) = \begin{cases}
0,&$if t is a root task$\cr
\max_{p \in Pred(t)} DL(p),&$otherwise.$\cr
\end{cases}
$$

Algorithm~\ref{alg:planworkflow} shows how SPSS creates low-cost plans for
each workflow. The \textsc{PlanWorkflow} procedure first calls 
\textsc{DeadlineDistribution} to assign sub-deadlines to tasks according
to the procedure described above. Then, the \textsc{PlanWorkflow} procedure
schedules tasks onto VMs, allocating new VMs when necessary. For each task 
in the workflow, the the least expensive slot is chosen to schedule the 
task so that it can be completed by  its deadline. VMs are allocated in 
blocks of one billing cycle (one hour) regardless of the size of the task. 
When computing the cost of scheduling a task on a given VM, the algorithm 
considers idle slots in blocks that were allocated for previous tasks to be 
free, while slots in new blocks cost the full price of the new block. For 
example, if a task has a runtime of 10 minutes, and the price of a one-hour 
billing cycle is \$1, then the algorithm will either schedule the task on 
an existing VM that has an idle slot larger than 10 minutes for a cost of \$0, 
or it will allocate a new block on an existing VM, or provision a new VM, for 
a cost of \$1. If the cost of slots on two different VMs is equal, then the 
slot with the earliest start time is chosen. To prevent too many VMs from 
being provisioned, the algorithm always prefers to extend the runtime of 
existing VMs before allocating new VMs. The result of this is that the 
algorithm will only allocate a new block if there are no idle slots on 
existing blocks large enough or early enough to complete the 
task by its deadline, and it will only allocate a new VM if it cannot add 
a block to an existing VM to complete the task by its deadline.

\begin{algorithm}
\caption{Workflow planning algorithm for SPSS}
\label{alg:planworkflow}
\begin{algorithmic}[1]
\Require $w$: workflow; $P$: current plan; $d$: deadline
\Ensure Create plan for $w$ that minimizes cost and meets deadline $d$
\Procedure{PlanWorkflow}{$w,P,d$}
    \State $P^\prime\gets$ copy of $P$
    \State \Call{DeadlineDistribution}{w,d}
    \For{$t\ in\ w\ sorted\ by\ DL(t)$}
        \State $v \gets$ VM that minimizes cost and start time of t
        \If{$FinishTime(t,v) < DL(t)$}
            \State Schedule(t,v)
        \Else
            \State Provision a new VM v
            \State Schedule(t,v)
        \EndIf
    \EndFor
    \State \textbf{return} $P^\prime$
\EndProcedure
\end{algorithmic}
\end{algorithm}

An example schedule generated by SPSS is shown in Fig.~\ref{fig:spss-example}.
This example shows how SPSS tends to start many workflows in parallel, running
each workflow over a longer period of time on only a few VMs to minimize cost. 
In comparison, the dynamic algorithms tend to run one workflow at a time across 
many VMs in parallel.
